Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-11 second-draft/pnas-word-order-si.tex
SI

\section{Formalization of Correlations}

@@

\begin{table*}[ht]
	\begin{center}
\small{
\begin{tabular}{|l|ll|l|l|}
	\hline
&	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{UD Relation}   & \multirow{2}{*}{\cite{greenberg1963universals}}    \\ 
&	verb & object & &   \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}&adposition    &    NP    &  $\xrightarrow{lifted\_case}$   & 3, 4   \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {2}}}&copula  verb  &    predicate    &    $\xrightarrow{lifted\_cop}$   & --    \\\hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}}&tense/aspect auxiliary    &    VP    &    \multirow{2}{*}{$\xleftarrow{aux}$}   & 16, 13  \\
&	negative auxiliary    &    VP    &    & -- \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}}&noun    &    genitive    &   $\xrightarrow{nmod}$ & 2, 23   \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}}&noun    &    relative clause    &    $\xrightarrow{acl}$ &  24    \\ \hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {6}}}}&complementizer    &    S    &   \multirow{2}{*}{$\xrightarrow{lifted\_mark}$}  & --    \\
&	adverbial subordinator & S &  & -- \\ \hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}}&	adjective & std. of comp. & \multirow{2}{*}{$\xrightarrow{obl}$} & --\\
&verb    &    PP    &    & 22   \\\hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}&`want'    &    VP    &    $\xrightarrow{xcomp}$   & 15   \\\hline
 \hline
\end{tabular}
}
	\end{center}
	\caption{Greenbergian Correlations based on Dryer \cite{dryer1992greenbergian}, with operationalizations with Universal Dependencies using the modified format of \cite{futrell2015largescale} (see text).
	For reference, we also provide the numbers of the closest corresponding universals stated in Greenberg's original study (if possible).
	}\label{table:greenberg-dryer}
\end{table*}

@@

\paragraph{z-Scored Data (Figure (REF))}
We z-transformed on the level of individual languages, normalizing the mean and SD parseability and predictability of the (1) real grammar, (2) the mean of predictability and parseability of all random grammars, (3) the grammar optimized for efficiency (at $\lambda =0.9$, see Section REF), (4) grammar optimized for parseability only, and (5) grammar optimized for predictability only.

Within each language, we estimate the Pareto curve by connecting the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section REF), (2) parseability, and (3) predictability.
We then estimated the overall average Pareto curve, as shown in Figure (REF), by averaging the z-transformed points (1), (2), (3).

@@

    \includegraphics[scale=.25]{../results/correlations/correlations-by-grammar/ground-corrs-efficiency.pdf}
    \includegraphics[scale=.25]{../results/correlations/correlations-by-grammar/ground-corrs-parseability.pdf}
    \includegraphics[scale=.25]{../results/correlations/correlations-by-grammar/ground-corrs-predictability.pdf}

@@

%To verify robustness to more informative priors, we conducted the analysis (REF) with a much tighter prior $\mathcal{N}(0,1)$.

%Note that an intercept $\alpha \approx 2.19$ in the logistic model corresponds to a correlation that holds in 90\% of cases; we thus selected the prior $\mathcal{N}(0,2)$.
%we obtained a qualitatively equivalent result (posterior probability $0.932$ of predicting all eight correlations).
% N(0,2): 0.9863
We obtained qualitatively equivalent results for analyses on the levels of individual relations when estimating frequentist models individually for each correlation using \texttt{lme4} \cite{bates2015fitting} and obtaining $p$-values from the resulting fit: For all eight correlations, the sign of $\alpha$ was positive ($p < 0.05$).

@@

Further, we report results for grammars optimized for dependency length minimization (DLM).
Following (CITE), we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars (CITE).
Indeed, we found that DLM is predicted by efficiency optimization, as reported in Discussion (Main Paper).
In line with these ideas, we show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.

@@

&nsubj  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_nsubj.pdf}  & See Section (REF) \\

@@

\end{center}
\caption{Predictions on UD relations with predictions from the typological literature. The first section contains the eight correlations discussed in the main paper (See Section REF); the second section provides other relations for which predictions are available. In the last column, we indicate what direction would be expected typologically.}
\label{tab:all-predictions-1}
\end{table*}

@@

The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
We previously carried out language optimization for $\lambda = 1$, and obtained essentially identical predictions for word order correlations.
We note that each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions (TODO check).

@@

The parameters of such a differentiable word order grammar are as follows. 
For each dependency label type $\tau$, we have (1) a \key{Direction Parameter} $a_\tau \in [0,1]$, and (2) a \key{Distance Parameter} $b_\tau \in \mathbb{R}$. 
Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p. 184) without replacement.

@@

A word order grammar can be viewed as a function $\mathcal{L}_\theta$, whose behavior is specified by parameters $\theta$, which takes an unordered dependency tree $t$ as input and produces as output an ordered sequence of words $\mathcal{L}_\theta(t) = \mathbf{w}$ linearizing the tree.
More generally, if $\mathcal{L}_\theta$ is a differentiable ordering grammar (Section~\ref{sec:diff-gramm}), then $\mathcal{L}_\theta(t)$ is a \emph{probability distribution} over ordered sequences of words $\mathbf{w}$.
In the limit where $\mathcal{L}_\theta$ becomes deterministic, the distribution $\mathcal{L}_\theta(t)$ concentrates on a single ordering.

Let $T$ be a distribution over unordered dependency trees and let $L_\theta$ be a differentiable word order grammar. Our goal is to find parameters $\theta$ which maximize an objective function $J$ in expectation over utterances:
\begin{equation}
\label{eq:exp-over-utts}
J_T(\theta) = \E_{t \sim T} \E_{\mathbf{w} \sim \mathcal{L}_\theta(t)} [R(t; \mathbf{w})], % TODO make the expectations look better!!
\end{equation}
where the function $R$ specifies a score for an individual utterance $\mathbf{w}$ expressing a tree $t$. In practice, $T$ will be the empirical distribution over unordered trees observed in a dependency treebank.
Below we discuss possible values of $R$, which may implement predictability, parseability, or a weighted combination of the two, representing overall efficiency.

We formalize predictability as follows. Given a language model $P_\phi$ with parameters $\phi$, the log probability of a sentence ${\bf w}$ is:
\begin{equation*}
R_{Pred}^\phi(t; {\bf w}) = \sum_{i=1}^{\#{\bf w}} \log P_\phi(w_{i+1}|w_{1\dots i}),
\end{equation*}
where the language model parameters $\phi$ are chosen predict the language induced by tree distribution $T$ and word order grammar $\mathcal{L}_\theta$ as accurately as possible (maximizing predictability):
\begin{equation*}
	\phi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R^\phi_{Pred}({\bf w}, \phi) ].
\end{equation*}

We specify parseability as the mutual information between strings and trees, or equivalently the quantity of information provided by strings about trees. Formally, we say the parseability score of an utterance $\mathbf{w}$ is the pointwise mutual information of the $\mathbf{w}$ and the tree $t$ it was generated from, calculated using a parser with parameters $\psi$:\footnote{Note that $P(t)$ is a constant term which will be ignored in the optimization process.}
\begin{align}
\nonumber
R_{Pars}^\psi(t; {\bf w}) &= \log \frac{P_\psi (t|\mathbf{w})}{P(t)} \\
\nonumber
&= \displaystyle \sum_{i=1}^{\#\mathbf{w}} \log P_\psi\left(\begin{array}{@{}l@{}}head(w_{i}) \\ label(w_i)\end{array}{\big |}\textbf{w}\right) - \log P(t),
\end{align}
where the parser parameters $\psi$ are chosen to maximize parsing accuracy:
\begin{equation*}
\psi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R_{Pars}({\bf w}, \psi) ].
\end{equation*}

For efficiency, we linearly combine the two terms:
\begin{equation}
\label{eq:combined-score}
R_{Efficiency}^{\phi,\psi}(t; {\bf w}) = R_{Pars}^\psi(t; {\bf w}) + \lambda R_{Pred}^\phi(t; {\bf w}),
\end{equation}

%\begin{equation}
%    \label{eq:depl}
%	R_{DepL}(t; {\bf w}) = -\sum_{i=1}^{\# {\bf w}} |i - head(t; w_i)|,
%\end{equation}
%where $head(t; i) \in \{1, ..., n\}$ is the index of the head of $w_i$, and the sum is negative because dependency length is seen as a cost.

For each of our objective functions, we seek parameters $\theta$ of a word order grammar $\mathcal{L}_\theta$ that maximize the average utility of sentences in the language obtained by linearizing unordered trees $t \sim T$ according to $\mathcal{L}_\theta$. The calculation of the predictability and parseability objectives also requires fitting language models parameterized by $\phi$ and parsers parameterized by $\psi$, respectively; and the calculation of the efficiency objective requires joint optimization of the word order grammar, language models, and parsers. The optimization problems come out to:
\begin{align}
%\max_\theta &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{DepL}(t; {\bf w}) \label{eq:maximize-depl}\\
\max_{\theta,\phi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pred}^{\phi(\theta)}(t; {\bf w}) \label{eq:maximize-pred}\\
\max_{\theta,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pars}^{\psi(\theta)}(t; {\bf w}) \label{Eq:maximize-pars} \\
\max_{\theta,\phi,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Efficiency}^{\psi(\theta), \phi(\theta)}(t; {\bf w}). \label{eq:maximize-efficiency} 
\end{align}
Note that the original word orders from the treebank never enter this objective---in particular, they do not influence the language model and the parser, which are entirely determined by $\theta$.

%The case of predictability, parseability, and efficiency are more challenging, as 
%e need to simultaneously look for an ordering model optimized for the loss of a language model and/or a parser, and for a language model and/or parser that are optimized for that given ordering model.
We solve Eqs.~\ref{eq:maximize-pred}--\ref{eq:maximize-efficiency} by performing stochastic gradient descent over all sets of parameters $\theta, \phi$, and/or $\psi$ simultaneously.
Below we discuss the process only for the predictability objective (Eq.~\ref{eq:maximize-pred}); the other objectives are optimized analogously.

We initialize all values $a_\tau := 0.5$, $b_\tau := 0$ (except for the \emph{obj} dependency, for which we fix $a_\tau$ to 0 or 1, see Section (REF)).

In each optimization step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim P_{L_\theta}(\cdot|t)$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator %(analogous for $R_{Pars}$ instead of $R_{Pred}$)
\begin{equation}\label{eq:estimator}
{ \partial_\theta \left(\log P_{L_\theta}({\bf w}|t)\right) \cdot R^\phi_{Pred} (t; {\bf w}) \choose  \partial_\phi R_{Pred}^\phi(t; {\bf w})}
%{- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot R_{Pred}(\bf{w}, \thetal) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}
\end{equation}
%$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
for the gradient
$${ \partial_\theta \choose \partial_\phi} \E_{{\bf w'} \sim L_\theta(t)} R_{Pred}^\phi(t; {\bf w'}).$$
This unbiased estimator is the ordinary gradient estimator for $\phi$, and the REINFORCE estimator for $\theta$ \citep{williams1992simple}.

See Section (REF) for the stopping criterion.

>>>>

\section{Formalization of Correlations}\label{sec:correlations}

@@

\begin{table*}[ht]
	\begin{center}
\small{
\begin{tabular}{|l|ll|l|l|}
	\hline
&	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{UD Relation}   & \multirow{2}{*}{Greenberg \cite{greenberg1963universals}}    \\ 
&	verb & object & &   \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}&adposition    &    NP    &  $\xrightarrow{lifted\_case}$   & 3, 4   \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {2}}}&copula  verb  &    predicate    &    $\xrightarrow{lifted\_cop}$   & --    \\\hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}}&tense/aspect auxiliary    &    VP    &    \multirow{2}{*}{$\xleftarrow{aux}$}   & 16, 13  \\
&	negative auxiliary    &    VP    &    & -- \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}}&noun    &    genitive    &   $\xrightarrow{nmod}$ & 2, 23   \\ \hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}}&noun    &    relative clause    &    $\xrightarrow{acl}$ &  24    \\ \hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {6}}}}&complementizer    &    S    &   \multirow{2}{*}{$\xrightarrow{lifted\_mark}$}  & --    \\
&	adverbial subordinator & S &  & -- \\ \hline
\multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}}&	adjective & std. of comp. & \multirow{2}{*}{$\xrightarrow{obl}$} & --\\
&verb    &    PP    &    & 22   \\\hline
\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}&`want'    &    VP    &    $\xrightarrow{xcomp}$   & 15   \\\hline
 \hline
\end{tabular}
}
	\end{center}
	\caption{Greenbergian Correlations based on Dryer \cite{dryer1992greenbergian}, with operationalizations with Universal Dependencies using the modified format of \cite{futrell2015largescale} (see text).
	For reference, we also provide the numbers of the closest corresponding universals stated in Greenberg's original study (if possible).
	}\label{table:greenberg-dryer}
\end{table*}

@@

\paragraph{z-Scored Data (Figure (REF))}
We z-transformed on the level of individual languages, normalizing the mean and SD parseability and predictability of the (1) real grammar, (2) the mean of predictability and parseability of all random grammars, (3) the grammar optimized for efficiency (at $\lambda =0.9$, see Section~\ref{sec:lambda}), (4) grammar optimized for parseability only, and (5) grammar optimized for predictability only.

Within each language, we estimate the Pareto curve by connecting the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section~\ref{sec:lambda}), (2) parseability, and (3) predictability.
We then estimated the overall average Pareto curve, as shown in Figure (REF), by averaging the z-transformed points (1), (2), (3).

@@

    \includegraphics[scale=.55]{../results/correlations/correlations-by-grammar/ground-corrs-efficiency.pdf}
    \includegraphics[scale=.55]{../results/correlations/correlations-by-grammar/ground-corrs-parseability.pdf}
    \includegraphics[scale=.55]{../results/correlations/correlations-by-grammar/ground-corrs-predictability.pdf}

@@

We obtained qualitatively equivalent results for analyses on the levels of individual relations when estimating frequentist models individually for each correlation using \texttt{lme4} \cite{bates2015fitting} and obtaining $p$-values from the resulting fit: For all eight correlations, the sign of $\alpha$ was positive ($p < 0.05$).

@@

\textcolor{red}{at some place also say the following: Dependency length quantified in this manner is a heuristic measure of complexity: the actual processing complexity induced by long dependencies is not a linear function of length and depends crucially on the types of dependencies involved \cite{demberg2008data} and the specific elements intervening between the head and dependent \cite{gibson1998linguistic,gibson2000dependency,lewis2005activationbased}. }

Further, we report results for grammars optimized for dependency length minimization (DLM).
Following \cite{futrell2015largescale}, we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
Prior work has suggested principles akin to DLM as approximating efficiency optimization of grammars~\cite{hawkins1994performance,futrell2017memory, futrell2017generalizing}.
Indeed, we found that DLM is predicted by efficiency optimization, as reported in Discussion (Main Paper).
In line with these ideas, we show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.

@@

&nsubj  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_nsubj.pdf}  & See Section~\ref{sec:correlations} \\

@@

\end{center}
\caption{Predictions on UD relations with predictions from the typological literature. The first section contains the eight correlations discussed in the main paper (See Section~\ref{sec:correlations}); the second section provides other relations for which predictions are available. In the last column, we indicate what direction would be expected typologically.}
\label{tab:all-predictions-1}
\end{table*}

@@

The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
We previously carried out language optimization for $\lambda = 1$, and obtained essentially identical predictions for word order correlations.
We note that each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions (Table~\ref{tab:all-predictions-1}).

@@

We extended the parameter space of grammars by continuously interpolating between grammars, making efficiency a \emph{differentiable} function of grammar parameters.
The parameters of such a \key{differentiable word order grammar} are as follows. 
For each dependency label type $\tau$, we have (1) a \key{Direction Parameter} $a_\tau \in [0,1]$, and (2) a \key{Distance Parameter} $b_\tau \in \mathbb{R}$. 
Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p. 184) without replacement.

@@

In this section, we describe how we optimized grammar parameters for efficiency.
A word order grammar can be viewed as a function $\mathcal{L}_\theta$, whose behavior is specified by parameters $\theta$, which takes an unordered dependency tree $t$ as input and produces as output an ordered sequence of words $\mathcal{L}_\theta(t) = \mathbf{w}$ linearizing the tree.
More generally, if $\mathcal{L}_\theta$ is a differentiable ordering grammar (Section~\ref{sec:diff-gramm}), then $\mathcal{L}_\theta(t)$ is a \emph{probability distribution} over ordered sequences of words $\mathbf{w}$.
In the limit where $\mathcal{L}_\theta$ becomes deterministic, the distribution $\mathcal{L}_\theta(t)$ concentrates on a single ordering.

\textcolor{red}{TODO I'll clean this up}

Recall the definition of efficiency
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where 
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
\begin{equation}
	R_{Pred} := - H[\mathcal{U}] = \sum_{u} p(u) \log p(u)
\end{equation}

These quatities are estimated using two neural models:
A \key{parser} recovers syntactic structures from utterances by computing a distribution $p_\phi(t|u)$, parameterized via parser parameters $\phi$.
A \key{language model} calculates the word-by-word surprisal of an utterance: $-\log p_\psi(u) = - \sum_{t=1}^{|u|} \log p_\psi(u_t|u_{1\dots t-1})$.
The parsers and language models appropriate to a tree distribution $\mathcal{T}$ and a differentiable ordering grammar $\mathcal{L}_\theta$ are the ones minimizing parsing loss and surprisal:
\begin{equation}
	\phi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R^\phi_{Pred}({\bf w}, \phi) ].
\end{equation}
\begin{equation}
\psi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R_{Pars}({\bf w}, \psi) ].
\end{equation}

Using these models, we can rewrite 
\begin{equation}
	R_{Pars}^{\psi(\theta)}= \E_t \E_{u \sim p_\theta(u|t)} R_{Pars}^{\psi(\theta)}(u,t) 
\end{equation}
where $R_{Pars}(u,t) = \log \frac{p_\phi(t|u)}{p(t)}$, and $p_\theta(u|t)$ is the probability assigned to sentence $u$ by the distribution $\mathcal{L}_\theta(t)$ defined by the differentiable word order grammar $\mathcal{L}_\theta$,
and
\begin{equation}
	R_{Pred} = \E_t \E_{u \sim p_\theta(u|t)} R_{Pred}(u)
\end{equation}
where $R_{Pred}(u) = \log p_\psi(u)$.
Therefore, we can rewrite efficiency as
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}}^\theta := \E_t \E_{u \sim p_\theta(u|t)} \left[R_{Pars}^{\phi(\theta)}(u,t) + \lambda R_{Pred}^{\psi(\theta)}(u)\right]
\end{equation}
Directly applying gradient descent to this expression is hard due to the complicated dependence of $\phi(\theta), \psi(\theta)$ on $\theta$.

However, we can rewrite
\begin{equation}\label{eq:efficiency}
\argmax_\theta\	R_{\textit{Eff}}^\theta 
\end{equation}
equivalently into
\begin{equation}\label{eq:efficiency}
\argmax_\theta\	\max_{\phi, \psi} R_{\textit{Eff}}^{\theta, \phi, \psi} = \argmax_\theta\	\max_{\phi, \psi} \E_t \E_{u \sim p_\theta(u|t)} \left[R_{Pars}^{\phi}(u,t) + \lambda R_{Pred}^{\psi}(u)\right]
\end{equation}
Importantly, $R_{\textit{Eff}}^{\theta, \phi, \psi}$ is differentiable in $\theta, \phi, \psi$: %, and we can apply stochastic gradient descent to carry out this optimization.

\begin{align}
\partial_\theta R_{\textit{Eff}}^{\theta, \phi, \psi} &= \E_t \E_{u \sim p_\theta(u|t)} \left[  \left[\partial_\theta \log p_\theta(u|t)\right] \cdot    \left(R_{Pars}^{\phi}(u,t) + \lambda R_{Pred}^{\psi}(u)\right) \right] \label{eq:dtheta}\\ 
\partial_\phi R_{\textit{Eff}}^{\theta, \phi, \psi} &= \E_t \E_{u \sim p_\theta(u|t)}  \left[\partial_\phi R_{Pars}^{\phi}(u,t)\right] \\
\partial_\psi R_{\textit{Eff}}^{\theta, \phi, \psi} &= \E_t \E_{u \sim p_\theta(u|t)}  \left[\lambda \cdot \partial_\psi R_{Pred}^{\psi}(u)\right] \label{eq:dpsi}
\end{align}
where \ref{eq:dtheta} is derived using the \emph{score-function} or \emph{REINFORCE} trick~\cite{williams1992simple}.

We can therefore apply stochastic gradient descent:
In each optimization step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim p_{\theta}(\cdot|t)$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator given by the expressions in the square brackets in \ref{eq:dtheta}-\ref{eq:dpsi}.

\textcolor{red}{fix which one of psi, phi goes with pred, parse}

Optimizing for only parseability (or predictability) is very similar -- in this case, the terms involving $R_{Pred}^\phi$ (or $R_{Pars}^\psi$) are removed.

At the beginning of the optimization procedure, we initialize all values $a_\tau := 0.5$, $b_\tau := 0$ (except for the \emph{obj} dependency, for which we fix $a_\tau$ to 0 or 1, see Section~\ref{sec:neural-architectures}).

See Section~\ref{sec:neural-architectures} for the stopping criterion and learning rates.

%
%%(analogous for $R_{Pars}$ instead of $R_{Pred}$)
%\begin{equation}\label{eq:estimator}
%{ \partial_\theta \left(\log P_{L_\theta}({\bf w}|t)\right) \cdot R^\phi_{Pred} (t; {\bf w}) \choose  \partial_\phi R_{Pred}^\phi(t; {\bf w})}
%%{- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot R_{Pred}(\bf{w}, \thetal) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}
%\end{equation}
%%$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
%for the gradient
%$${ \partial_\theta \choose \partial_\phi} \E_{{\bf w'} \sim L_\theta(t)} R_{Pred}^\phi(t; {\bf w'}).$$
%This unbiased estimator is the ordinary gradient estimator for $\phi$, and the REINFORCE estimator for $\theta$ \citep{williams1992simple}.
%

%
%
%Let $T$ be a distribution over unordered dependency trees and let $L_\theta$ be a differentiable word order grammar. Our goal is to find parameters $\theta$ which maximize an objective function $J$ in expectation over utterances:
%\begin{equation}
%\label{eq:exp-over-utts}
%J_T(\theta) = \E_{t \sim T} \E_{\mathbf{w} \sim \mathcal{L}_\theta(t)} [R(t; \mathbf{w})], % TODO make the expectations look better!!
%\end{equation}
%where the function $R$ specifies a score for an individual utterance $\mathbf{w}$ expressing a tree $t$. In practice, $T$ will be the empirical distribution over unordered trees observed in a dependency treebank.
%Below we discuss possible values of $R$, which may implement predictability, parseability, or a weighted combination of the two, representing overall efficiency.
%
%
%
%We formalize predictability as follows. Given a language model $P_\phi$ with parameters $\phi$, the log probability of a sentence ${\bf w}$ is:
%\begin{equation*}
%R_{Pred}^\phi(t; {\bf w}) = \sum_{i=1}^{\#{\bf w}} \log P_\phi(w_{i+1}|w_{1\dots i}),
%\end{equation*}
%where the language model parameters $\phi$ are chosen predict the language induced by tree distribution $T$ and word order grammar $\mathcal{L}_\theta$ as accurately as possible (maximizing predictability):
%\begin{equation*}
%	\phi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R^\phi_{Pred}({\bf w}, \phi) ].
%\end{equation*}
%
%
%We specify parseability as the mutual information between strings and trees, or equivalently the quantity of information provided by strings about trees. Formally, we say the parseability score of an utterance $\mathbf{w}$ is the pointwise mutual information of the $\mathbf{w}$ and the tree $t$ it was generated from, calculated using a parser with parameters $\psi$:\footnote{Note that $P(t)$ is a constant term which will be ignored in the optimization process.}
%\begin{align}
%\nonumber
%R_{Pars}^\psi(t; {\bf w}) &= \log \frac{P_\psi (t|\mathbf{w})}{P(t)} \\
%\nonumber
%&= \displaystyle \sum_{i=1}^{\#\mathbf{w}} \log P_\psi\left(\begin{array}{@{}l@{}}head(w_{i}) \\ label(w_i)\end{array}{\big |}\textbf{w}\right) - \log P(t),
%\end{align}
%where the parser parameters $\psi$ are chosen to maximize parsing accuracy:
%\begin{equation*}
%\psi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R_{Pars}({\bf w}, \psi) ].
%\end{equation*}
%
%For efficiency, we linearly combine the two terms:
%\begin{equation}
%\label{eq:combined-score}
%R_{Efficiency}^{\phi,\psi}(t; {\bf w}) = R_{Pars}^\psi(t; {\bf w}) + \lambda R_{Pred}^\phi(t; {\bf w}),
%\end{equation}
%
%
%%\begin{equation}
%%    \label{eq:depl}
%%	R_{DepL}(t; {\bf w}) = -\sum_{i=1}^{\# {\bf w}} |i - head(t; w_i)|,
%%\end{equation}
%%where $head(t; i) \in \{1, ..., n\}$ is the index of the head of $w_i$, and the sum is negative because dependency length is seen as a cost. 
%
%For each of our objective functions, we seek parameters $\theta$ of a word order grammar $\mathcal{L}_\theta$ that maximize the average utility of sentences in the language obtained by linearizing unordered trees $t \sim T$ according to $\mathcal{L}_\theta$. The calculation of the predictability and parseability objectives also requires fitting language models parameterized by $\phi$ and parsers parameterized by $\psi$, respectively; and the calculation of the efficiency objective requires joint optimization of the word order grammar, language models, and parsers. The optimization problems come out to:
%\begin{align}
%%\max_\theta &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{DepL}(t; {\bf w}) \label{eq:maximize-depl}\\
%\max_{\theta,\phi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pred}^{\phi(\theta)}(t; {\bf w}) \label{eq:maximize-pred}\\
%\max_{\theta,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Pars}^{\psi(\theta)}(t; {\bf w}) \label{Eq:maximize-pars} \\
%\max_{\theta,\phi,\psi} &\E_{t \sim T}\E_{{\bf w} \sim L_\theta(t)} R_{Efficiency}^{\psi(\theta), \phi(\theta)}(t; {\bf w}). \label{eq:maximize-efficiency} 
%\end{align}
%Note that the original word orders from the treebank never enter this objective---in particular, they do not influence the language model and the parser, which are entirely determined by $\theta$.
%
%
%%The case of predictability, parseability, and efficiency are more challenging, as 
%%e need to simultaneously look for an ordering model optimized for the loss of a language model and/or a parser, and for a language model and/or parser that are optimized for that given ordering model.
%We solve Eqs.~\ref{eq:maximize-pred}--\ref{eq:maximize-efficiency} by performing stochastic gradient descent over all sets of parameters $\theta, \phi$, and/or $\psi$ simultaneously.
%Below we discuss the process only for the predictability objective (Eq.~\ref{eq:maximize-pred}); the other objectives are optimized analogously.
%
%We initialize all values $a_\tau := 0.5$, $b_\tau := 0$ (except for the \emph{obj} dependency, for which we fix $a_\tau$ to 0 or 1, see Section (REF)).
%
%In each optimization step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim P_{L_\theta}(\cdot|t)$.
%Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
%do a gradient descent step using the estimator %(analogous for $R_{Pars}$ instead of $R_{Pred}$)
%\begin{equation}\label{eq:estimator}
%{ \partial_\theta \left(\log P_{L_\theta}({\bf w}|t)\right) \cdot R^\phi_{Pred} (t; {\bf w}) \choose  \partial_\phi R_{Pred}^\phi(t; {\bf w})}
